Data Exploration, Quality Control, and Statistical Analysis of ChIP-exo/nexus Experiments

### 2016-07-06 paper/chipexo_QC_vs2.tex
corrected first part of comments, until half of results section

\begin{fmbox}
\dochead{Draft}

@@

\author[
   addressref={aff1},
   noteref={n1},
   email={welch@stat.wisc.edu}      
]{\inits{RW}\fnm{Rene} \snm{Welch} }
\author[
   addressref={aff6},        % id's of addresses, e.g. {aff1,aff2}
   noteref={n1},
   email={chungd@musc.edu}   % email address
]{\inits{DC}\fnm{Dongjun} \snm{Chung} } 
\author[
   addressref={aff3},
   email={ong@cs.wisc.edu}
]{\inits{IO}\fnm{Irene} \snm{Ong}}
\author[
   addressref={aff3,aff4},
   email={jagrass@wisc.edu}
]{\inits{JG}\fnm{Jeffrey} \snm{Grass}}
\author[
   addressref={aff3,aff4,aff5},
   email={landick@bact.wisc.edu}
]{\inits{RL}\fnm{Robert} \snm{Landick}}
\author[
   addressref={aff1,aff2},
   corref={aff1},
   email={keles@stat.wisc.edu}
]{\inits{SK}\fnm{S\"und\"uz} \snm{Kele\c{s}}}

@@

  \begin{abstract}
    ChIP-exo/nexus experiments present modifications on the commonly
    used ChIP-seq protocol for high resolution mapping of
    transcription factor binding sites. Although many aspects of the
    ChIP-exo data analysis are similar to those of ChIP-seq, these
    high throughput experiments present a number of unique quality
    control and analysis challenges. We develop a statistical quality
    control pipeline and accompanying \texttt{R} package, \pname{}, to
    enable exploration and analysis of ChIP-exo and related
    experiments. \pname{} evaluates a number of key issues including
    strand imbalance, library complexity, and signal enrichment of
    data from ChIP-exo/nexus experiments. Assessment of these
    properties are facilitated through diagnostic plots and summary
    statistics calculated over regions of the genome with varying
    levels of coverage.
    %\pname pipeline explores and quantifies these aspects by partitioning the experiment reads into a collection of regions, calculating a series of summary statistics for each region, providing visualizations and calculating measures to globally asses t quality of a ChIP-exo experiment.

    We evaluated our QC pipeline with both large collections of public
    ChIP-exo/nexus data and multiple, new ChIP-exo datasets from
    \textit{E. Coli}.  \pname{} analysis of these datasets resulted in
    guidelines for using these QC metrics across a wide range of
    sequencing depths and further insights for modeling ChIP-exo
    data. %In addition, we demonstrate that the
    %ChIP-exo QC pipeline it is also applicable to ChIP-nexus data,
    %showing that those experiments present higher quality than
    %ChIP-exo experiments under similar conditions.
    Finally, although ChIP-exo experiments have been compared to
    ChIP-seq experiments with single-end (SE) sequencing, we provide,
    for the first time, comparisons with paired-end (PE) ChIP-seq
    experiments.
    %We compared ChIP-exo with Paired End (PE) and Single End (SE)
    %ChIP-Seq and found the following characteristics: First, although
    %often assumed in ChIP-exo data analysis methods, the ``peak pair''
    %assumptions does not hold locally in actual ChIP-exo data. Second,
    %we for the first time compared PE ChIP-Seq with ChIP-exo and found
%    that 
    We illustrate that, at fixed sequencing depths, ChIP-exo provides
    higher sensitivity, specificity, and spatial resolution than PE
    ChIP-seq and both significantly outperform their SE ChIP-seq
    counterpart. Furthermore, we show that ChIP-exo and PE ChIP-seq
    are comparable in sensitivity for closely located binding events,
    but as the average distance between binding events increases,
    ChIP-exo exhibits higher sensitivity than PE ChIP-Seq.

    \SK{The last sentence does not really make much sense? }, \RW{I
      meant that when we have a peak composed by more than one binding
      event, as the mean distance between binding events increases
      then ChIP-exo is more likely to identify all the binding events,
      I used sensitivity because we defined it as the proportion of
      identified binding events in a given peak. I think it depends if
      we decide to add Fig8 in the main article or in the
      supplement.}

  \end{abstract}

\begin{keyword}
  \kwd{ChIP-exo}
  \kwd{ChIP-nexus}
  \kwd{ChIP-seq}
  \kwd{Statistical Quality Control}
  \kwd{Spatial Resolution}
  \kwd{Transcription Factor}  
  \kwd{Binding Site Identification with High-Resolution}
  \kwd{Deconvolution}
\end{keyword}

@@

Chromatin Immunoprecipitation followed by exonuclease digestion and
next generation sequencing (ChIP-exo) is currently one of the
state-of-the-art high throughput assays for profiling protein-DNA
interactions at or close to single base-pair resolution
\cite{exo1}. It presents a powerful alternative to popularly used
ChIP-seq (Chromatin immunoprecipitation coupled with next generation
sequencing) assay. ChIP-exo experiments first capture millions of DNA
fragments ($150$ - $250$ bps in length) that the protein under study
interacts with using a protein-specific antibody and random
fragmentation of DNA. Then, $\lambda$-exonuclease ($\lambda$-exo) is
deployed to trim the $5^{\prime}$ end of each DNA fragment to each
protein-DNA interaction boundary. This step is unique to ChIP-exo and
aims to achieve significantly higher spatial resolution compared to
ChIP-seq. Finally, high throughput sequencing of a small region ($36$
to $100$ bps) at the $5^{\prime}$ end of each fragment generates
millions of reads.  Similarly, ChIP-nexus (Chromatin
Immunoprecipitation followed by exonuclease digestion, unique barcode,
single ligation and next generation ligation) \cite{chipnexus} is a
further modification on the ChIP-exo protocol. ChIP-nexus aims to
overcome limitations of ChIP-exo by yielding high complexity libraries
with numbers of cells comparable to that of ChIP-seq experiments. This
is achieved by reducing the numbers of ligations in the standard
ChIP-exo protocol from two to one, and adding unique, randomized
barcodes to adaptors to enable monitoring of
overamplification. Figure~\ref{fig:1}A illustrates the differences between
different ChIP-based protocols: ChIP-exo, single-end (SE) ChIP-seq,
paired-end (PE) ChIP-seq, ChIP-nexus. The $5^{\prime}$ ends of a
ChIP-exo/nexus experiment are clustered more tightly around the
binding sites of the protein than in a ChIP-seq experiment. In a PE
ChIP-seq experiment, both ends are sequenced as opposed to only the
$5^{\prime}$ end in a SE ChIP-seq.

% \RW{This modification makes a library generated
%   using the ChIP-nexus protocol to be larger than one generated with
%   ChIP-exo, which }
%where both sequencing adaptors are ligated at the
%end of the ChIP fragments. Then, after exonuclease digestion, DNA self-circularization with circLigase, and restriction enzyme cutting between the two adaptors, the final library is amplified. Compared to ChIP-exo, ChIP-nexus aims to   attain  higher resolution analysis while yielding higher complexity libraries.

\SK{Note to Rene: ChIP-nexus paper has a good description of what can
  go wrong with ChIP-exo; It seems like most of the read imbalance
  could also be due to ligation inefficiency. Although this probably
  does not explain why we have reads only from one strand in some
  regions. This is more likely explained by over digestion in one
  strand or single stranded TF-DNA interaction. Think a bit more on
  these issues and understand how the ligation could give rise to
  strand imbalance, we can ask Bob to have a closer look at the
  section where we have this discussion.}

% \RW{1 - I think it may come from the starting material in each side. I
%   think that at least the number of fragments on each side must
%   preserve that if there is more starting reads in one side, then this
%   must be the case after the amplification.}

% \RW{2 - There is not control about how the enzyme digests the
%   fragments, as far as we know it starts on the $5\prime$ side and
%   stops until it finds the TF or it runs out of energy to keep
%   digesting. I think Bob mentioned that it may be possible for the
%   enzyme to pass though the TF if it is left digesting for a longer
%   time, but if that's the case then the question may be why it
%   preffers one strand over the other, which may lead to the GC-bias
%   but as far as I can tell is very hard to observe that because that
%   bias appear also in ChIP-seq (and it seeems that also appear in the
%   circularization step from ChIP-nexus).}

% \RW{3 - I think that the regions formed by fragments in one strand,
%   may be copies of few undigested fragments, in a good experiment the
%   fragments are expected to be clustered around the binding sites,
%   therefore it is less likely to observe this fragments scattered in
%   the genome. In a bad experiment is the opposite case, so the
%   fragments are not clustered so if the enzyme doesn't have a strong
%   preference to how is attracted to the fragments then the remaining
%   ones can be seen as random positions around the genome, this
%   positions are copied several times during the amplification and that
%   may be the reason why we observe something as scattered. Perhaps if
%   the enzyme was left for longer time, then the starting material that
%   later was amplified into this regions would have been digested and
%   some of this regions wouldn't appear (I am thinking about the deeply
%   sequenced TBP experiments.)}

@@

To address these challenges, we develop a suite of diagnostic plots
and summary statistics and implement them in a versatile \texttt{R
  package} named \pname{}.  We apply this pipeline on a large
collection of public and newly generated ChIP-exo/nexus data. We
validate implications of the QC pipeline by evaluation of the samples
for features that capture high signal to noise such as occurrences of
motifs recognized by the profiled DNA interacting protein. Our
analyses of this large collection of data revealed that the so-called
ChIP-exo peak-pair assumption is subject to violations. To further
address this and provide a platform where ChIP-exo and ChIP-seq
experiments can be evaluated with comparable methods, we assess
performances of recently developed methods suitable for ChIP-exo
analysis, including dpeak \cite{dpeak} and GEM \cite{gem}. We observe
that dPeak performs as good or better than the available ChIP-exo
methods and provides a platform where PE and SE ChIP-seq can be
compared with their ChIP-exo counterpart. Our comparisons of PE
ChIP-seq with ChIP-exo interestingly highlights that while ChIP-exo
outperforms PE ChIP-seq in terms of resolution and detection power,
both are significantly better than SE ChIP-seq.

% while there are not established QC pipelines for ChIP-exo; previous
% ChIP-exo analyses used ChIP-Seq samples to compare the resolution
% between experiments \cite{exo1,exoillumina,exo2}.

%to interrogate these
%biases in a ChIP-exo experiment and globally asses the enrichment and
%library complexity of a ChIP-exo sample and a procedure to distinguish
%low complexity libraries from deeply sequenced experiments. 
%This
%aspect is unique to ChIP-exo, since the exonuclease enzyme it is
%expected to digested the reads that are not bound to a transcription
%factor, therefore the number of bases where a ChIP-exo fragment could
%be potentially aligned is reduced. That way, in a high quality
%ChIP-exo experiment is possible to observe a large amount of reads to
%be aligned to unique position due to genomic signal instead of PCR
%artifacts.

%In order to obtain the potential benefits of ChIP-exo on protein
%binding site identification, it is critical to use algorithms that
%could fully utilize information available in ChIP-exo data. Rhee and
%Pugh, 2011 \nocite{exo1} discussed that reads in the forward and reverse
%strand might construct peak pairs around bound proteins, of which
%heights were implicitly assumed to be symmetric. Based on this
%rationale, they used the ``peak pair method'' that predicts the
%midpoint of two modes of peak pairs as potential binding
%sites. Recently developed ChIP-exo data analysis methods, such as Mace
%(Wang et al, 20114\nocite{mace}), CexoR (Madrigal, 2015\nocite{cexor})
%and Peakzilla (Bardet et al., 2013\nocite{peakzilla}), are also based
%on this peak pair assumption. However, appropriateness of such
%assumption was not fully evaluated in the literature yet. Furthermore,
%it is still unknown which factors could affect protein binding site
%identification using ChIP-exo data. In order to address this problem,
%we investigated various aspects of ChIP-exo data by contrasting them
%with their respective ChIP-Seq experiments.

%Currently, research on statistical methods for ChIP-exo data is still
%in its very early stage. Although many methods have been proposed to
%identify protein binding sites from ChIP-Seq data (reviewed by
%Wilbanks and Facciotti, 2012\nocite{evaluation} and Pepke and Wold,
%2009 \nocite{computation}), such as MACS (Zhang et al.,
%2008\nocite{macs}), CisGenome (Ji et al., 2008\nocite{cisgenome}) and
%MOSAiCS (Kuan et al., 2009\nocite{mosaics}), these approaches might
%not fully utilize potentials of ChIP-exo data for high resolution
%identification of protein binding sites. Specifically these approaches
%reveal protein binding sites only in lower resolution, i.e., at an
%interval of hundreds to thousands of base pairs. Furthermore, they
%implicitly assume that there is only one ``mode'' or ``predicted
%binding location'' per this wide genomic interval. More recently,
%deconvolution algorithms such as Deconvolution (Lun et al.,
%2009\cite{csdeconv}), GEM (Guo et al., 2012\nocite{gem}, an improved
%version of Guo et al., 2010\nocite{gps} ) and PICS (Zhang et al.,
%2010\nocite{pics}) have been proposed to identify binding sites in
%higher resolution using ChIP-Seq data. However, most of them are still
%not tailored for ChIP-exo and PE and SE ChIP-Seq data in a unified
%framework and as a result, currently available methods are not
%appropriate for fair comparison between ChIP-exo and ChIP-Seq. To
%address these limitations, we developed and utilized an improved
%version of dPeak (Chung et al., 2013\nocite{dpeak}), a high resolution
%binding site identification (deconvolution) algorithm that we
%previously developed for PE and SE ChIP-Seq data, so that it can also
%handle ChIP-exo data. The dPeak algorithm implements a probabilistic
%model that accurately describes the ChIP-exo and ChIP-Seq data
%generation process.

%Some of the key findings in this work are as follows. First, we
%demonstrate that the ``peak pair'' assumption of Rhee and Pugh,
%2013\nocite{exo2} does not hold well in real ChIP-exo data. Second, we
%found that when analyzing ChIP-exo data and the Input control is not
%available, it is useful to adjust for GC content and mappability
%biases to improve peak calling and binding site identification. Third,
%we evaluated several methods to identify binding events and dPeak
%performs competitively respect to GEM and MACE when analyzing ChIP-exo
%data. Finally, when comparable number of reads is used for both
%ChIP-exo and ChIP-Seq , dPeak coupled with ChIP-exo data provides
%resolution comparable to PE ChIP-Seq and both significantly improve
%the resolution of protein binding identification compared to SE-based
%analysis with any of the available methods.

@@

\subsection*{Publicly available ChIP-exo/nexus and novel \textit{E. coli} ChIP-seq/exo datasets}
We utilized a rich collection of publicly available ChIP-exo/nexus
data from multiple organisms to build and evaluate our quality control
pipeline (Table~\ref{tab:qc}). These include: CTCF factor in human
HeLa cells \cite{exo1}; ER factor in human MCF-7 cells
\cite{exoillumina}; GR factor in IMR90, K562, and U2OS human cells
\cite{starick15}; TBP factor in human K562 cells \cite{venters13}.
ChIP-nexus data included experiments from \cite{chipnexus} profiling
TBP in human K562 cells, MyC and Max in \textit{D. melanogaster} S2
cells, and Twist and Dorsal in \textit{D. melanogaster} embryo.

In order to have a setting where we can compare SE and PE ChIP-seq
with their ChIP-exo counterpart, we profiled $\sig$ under a variety of
conditions in \textit{E. coli} with both ChIP-exo and PE ChIP-seq
(Table~\ref{tab:qc_sig}). Collectively, we generated $\sig$ factor
ChIP-exo, PE and SE ChIP-seq experiments under aerobic ($+O_2$) and
anaerobic ($-O_2$) conditions in glucose minimal media, for simplicity
we named these experiments as E1, P1 and S1 respectively. Similarly,
we generated $\sig$ factor ChIP-exo, PE and SE (generated \textit{in
  silico}) ChIP-seq experiments in \textit{E. coli} under aerobic
($+O_2$) conditions with and without rifampicin treatment.  We refer
to these latter set of experiments as time point 20 minutes and time
point 0 minutes, we also named these experiments E2, P2 and S2 for
ChIP-exo, PE ChIP-seq and SE ChIP-seq respectively.
% \SK{Check for
%   consistency with other parts of the paper and correct, ok to use
%   different naming convention}.
We specifically used these experimental data for comparisons of
ChIP-exo and PE ChIP-seq assays in identifying closely spaced binding
events and their resolution, i.e., physical proximity of the predicted
events to the actual binding sites.  This comparison benefits from
using $\sig$ which is a transcription initiation factor of
housekeeping genes in \textit{E. coli}. Many \textit{E. coli}
promoters contain multiple transcription start sites (TSS). These TSSs
are often closely spaced, i.e., within 10 $\sim$ 150 bps of each
other, and are considered to be multiple ``switches'' that
differentially regulate gene expression under diverse growth
conditions \cite{regulondb}.

@@

%We continued our exploration by investigating whether the current
%state-of-the-art QC pipelines for ChIP-Seq are suitable for ChIP-exo
%and ChIP-nexus. In Tables \ref{tab:qc_sig} and \ref{tab:qc} we
%calculated a collection of the commonly used ChIP-Seq QC metrics using
%the ChIP-exo and ChIP-nexus experiments instead: Normalized Strand
%Cross-Correlation (NSC), Relative Strand Cross-Correlation and PCR
%Bottleneck Coefficient (PBC) defined as in
%\url{https://genome.ucsc.edu/ENCODE/qualityMetrics.html}\nocite{encode_qc}.

@@

\subsubsection*{Saturation analysis with ChIP-exo and ChIP-seq: Systematic comparison of ChIP-seq vs ChIP-exo under
  varying sequencing depths}
Establishing competitive performance of dPeak compared to other
ChIP-exo analysis methods enable comparison of ChIP-exo and ChIP-seq
with a unified analysis framework using dPeak. Previous comparisons of
ChIP-exo and SE ChIP-seq were all performed without controlling for
the sequencing depths of the samples. More importantly, although it is
well established that PE ChIP-seq leads to better resolution that SE
ChIP-seq in terms of binding site identification \cite{zhang16},
previous work does not have an in depth comparison of ChIP-exo and PE
ChIP-seq.
%Previously, ChIP-exo and SE ChIP-Seq have been compared only at a
%fixed depth level in the literature, while they did not include PE
%ChIP-Seq as well either. 
In order to address these limitations of the previous studies, we
performed a sub-sampling experiment by sampling fixed numbers of reads
from each of the $\sig$ ChIP-exo, PE ChIP-seq, and SE ChIP-seq
datasets. Specifically, for every $N$ reads in ChIP-exo and SE
ChIP-seq, $N/2$ read pairs were sampled for PE ChIP-seq to operate
under fixed sequencing costs.
%Additionally, it is worth noticing that for PE ChIP-Seq we
%sampled both ends of the fragment, hence for each sequencing depth we
%are sampling the half amount of pairs for PE ChIP-Seq than for
%ChIP-exo or SE ChIP-Seq.

>>>>

\begin{fmbox}
\dochead{Research}

@@

\author[
   addressref={aff1},
   noteref={n1},
   email={welch@stat.wisc.edu}      
]{\inits{RW}\fnm{Rene} \snm{Welch} }
\author[
   addressref={aff6},        % id's of addresses, e.g. {aff1,aff2}
   noteref={n1},
   email={chungd@musc.edu}   % email address
]{\inits{DC}\fnm{Dongjun} \snm{Chung} } 
\author[
   addressref={aff3,aff4},
   email={jagrass@wisc.edu}
]{\inits{JG}\fnm{Jeffrey} \snm{Grass}}
\author[
   addressref={aff3,aff4,aff5},
   email={landick@bact.wisc.edu}
]{\inits{RL}\fnm{Robert} \snm{Landick}}
\author[
   addressref={aff1,aff2},
   corref={aff1},
   email={keles@stat.wisc.edu}
]{\inits{SK}\fnm{S\"und\"uz} \snm{Kele\c{s}}}

@@

\begin{abstract}
ChIP-exo/nexus experiments present modifications on the commonly used ChIP-seq protocol for high resolution mapping of transcription factor binding sites. Although many aspects of the ChIP-exo data analysis are similar to those of ChIP-seq, these high throughput experiments pose a number of unique quality control and analysis challenges. We develop a statistical quality control pipeline and accompanying \texttt{R} package, \pname{}, to enable exploration and analysis of ChIP-exo and related experiments. \pname{} evaluates a number of key issues including strand imbalance, library complexity, and signal enrichment of data. Assessment of these features are facilitated through diagnostic plots and summary statistics calculated over regions of the genome with varying levels of coverage.

We evaluated our QC pipeline with both large collections of public ChIP-exo/nexus data and multiple, new ChIP-exo datasets from \textit{E. Coli}. \pname{} analysis of these datasets resulted in guidelines for using these QC metrics across a wide range of sequencing depths and further insights for modelling ChIP-exo data.Finally, although ChIP-exo experiments have been compared to ChIP-seq experiments with single-end (SE) sequencing, we provide, for the first time, comparisons with paired-end (PE) ChIP-seq experiments. We illustrate that, at fixed sequencing depths, ChIP-exo provides higher sensitivity, specificity, and spatial resolution than PE ChIP-seq and both significantly outperform their SE ChIP-seq counterpart. Furthermore, we show that for binding events separated by 200-400 bp ChIP-exo exhibits a significantly higher sensitivity.

%    \SK{The last sentence does not really make much sense? }, \RW{I
%      meant that when we have a peak composed by more than one binding
%      event, as the mean distance between binding events increases
%      then ChIP-exo is more likely to identify all the binding events,
%      I used sensitivity because we defined it as the proportion of
%      identified binding events in a given peak. I think it depends if
%      we decide to add Fig8 in the main article or in the
%      supplement.}

%\pname pipeline explores and quantifies these aspects by partitioning the experiment reads into a collection of regions, calculating a series of summary statistics for each region, providing visualizations and calculating measures to globally asses t quality of a ChIP-exo experiment. 
%In addition, we demonstrate that the ChIP-exo QC pipeline it is also applicable to ChIP-nexus data, showing that those experiments present higher quality than ChIP-exo experiments under similar conditions.
%We compared ChIP-exo with Paired End (PE) and Single End (SE) ChIP-Seq and found the following characteristics: First, although often assumed in ChIP-exo data analysis methods, the ``peak pair'' assumptions does not hold locally in actual ChIP-exo data. Second, we for the first time compared PE ChIP-Seq with ChIP-exo and found that

% ChIP-exo and PE ChIP-seq are comparable in sensitivity for closely located binding events, but as the average distance between binding events increases, ChIP-exo exhibits higher sensitivity than PE ChIP-seq.      
\end{abstract}

\begin{keyword}
\kwd{ChIP-exo}
\kwd{ChIP-nexus}
\kwd{ChIP-seq}
\kwd{Statistical Quality Control}
\kwd{Spatial Resolution}
\kwd{Transcription Factor}  
\kwd{Deconvolution}
\kwd{Binding Site Identification with High-Resolution.}
\end{keyword}

@@

Chromatin Immunoprecipitation followed by exonuclease digestion and next generation sequencing (ChIP-exo) is currently one of the state-of-the-art high throughput assays for profiling protein-DNA interactions at or close to single base-pair resolution \cite{exo1}. It presents a powerful alternative to popular ChIP-seq (Chromatin Immunoprecipitation coupled with next generation sequencing) assay. ChIP-exo experiments first capture millions of DNA fragments ($150$ - $250$ bps in length) that the protein under study interacts with using a protein-specific antibody and random fragmentation of DNA. Then, $\lambda$-exonuclease ($\lambda$-exo) is deployed to trim the $5^{\prime}$ end of each DNA fragment to each protein-DNA interaction boundary. This step is unique to ChIP-exo and aims to achieve significantly higher spatial resolution compared to ChIP-seq. Finally, high throughput sequencing of a small region ($36$ to $100$ bps) at the $5^{\prime}$ end of each fragment generates millions of reads. Similarly, ChIP-nexus (Chromatin Immunoprecipitation followed by exonuclease digestion, unique barcode, single ligation and next generation ligation) \cite{chipnexus} is a further modification on the ChIP-exo protocol. ChIP-nexus aims to overcome limitations of ChIP-exo by yielding high complexity libraries with numbers of cells comparable to that of ChIP-seq experiments. This is achieved by reducing the numbers of ligations in the standard ChIP-exo protocol from two to one, and adding unique, randomized barcodes to adaptors to enable monitoring of overamplification. Figure~\ref{fig:1}A illustrates the differences between different ChIP-based protocols: ChIP-exo, single-end (SE) ChIP-seq, paired-end (PE) ChIP-seq, ChIP-nexus. The $5^{\prime}$ ends of a ChIP-exo/nexus experiment are clustered more tightly around the binding sites of the protein than in a ChIP-seq experiment. In a PE ChIP-seq experiment, both ends are sequenced as opposed to only the $5^{\prime}$ end in a SE ChIP-seq.

% \RW{This modification makes a library generated using the ChIP-nexus protocol to be larger than one generated with ChIP-exo, which where both sequencing adaptors are ligated at the end of the ChIP fragments. Then, after exonuclease digestion, DNA self-circularization with circLigase, and restriction enzyme cutting between the two adaptors, the final library is amplified. Compared to ChIP-exo, ChIP-nexus aims to   attain  higher resolution analysis while yielding higher complexity libraries.

% \SK{Note to Rene: ChIP-nexus paper has a good description of what can go wrong with ChIP-exo; It seems like most of the read imbalance could also be due to ligation inefficiency. Although this probably does not explain why we have reads only from one strand in some regions. This is more likely explained by over digestion in one strand or single stranded TF-DNA interaction. Think a bit more on these issues and understand how the ligation could give rise to strand imbalance, we can ask Bob to have a closer look at the section where we have this discussion.}

% \RW{1 - I think it may come from the starting material in each side. I think that at least the number of fragments on each side must preserve that if there is more starting reads in one side, then this must be the case after the amplification.} 
% \RW{2 - There is not control about how the enzyme digests the fragments, as far as we know it starts on the $5\prime$ side and stops until it finds the TF or it runs out of energy to keep digesting. I think Bob mentioned that it may be possible for the enzyme to pass though the TF if it is left digesting for a longer time, but if that's the case then the question may be why it preffers one strand over the other, which may lead to the GC-bias but as far as I can tell is very hard to observe that because that bias appear also in ChIP-seq (and it seeems that also appear in the circularization step from ChIP-nexus).}

% \RW{3 - I think that the regions formed by fragments in one strand, may be copies of few undigested fragments, in a good experiment the    fragments are expected to be clustered around the binding sites, therefore it is less likely to observe this fragments scattered in the genome. In a bad experiment is the opposite case, so the fragments are not clustered so if the enzyme doesn't have a strong preference to how is attracted to the fragments then the remaining ones can be seen as random positions around the genome, this positions are copied several times during the amplification and that may be the reason why we observe something as scattered. Perhaps if the enzyme was left for longer time, then the starting material that later was amplified into this regions would have been digested and some of this regions wouldn't appear (I am thinking about the deeply sequenced TBP experiments.)}

@@

To address these challenges, we develop a suite of diagnostic plots and summary statistics and implement them in a versatile \texttt{R ackage} named \pname{}.  We apply this pipeline on a large collection of public and newly generated ChIP-exo/nexus data. We validate the QC pipeline by evaluation of the samples for features that capture high signal to noise such as occurrences of motifs recognized by the profiled DNA interacting protein. Our analyses of this large collection of data reveal that the ChIP-exo peak-pair assumption, underlying most of the ChIP-exo analysis pipelines is subject to violations. To further address this and provide a platform where ChIP-exo and ChIP-seq experiments can be evaluated with comparable methods, we assess performances of recently developed methods suitable for ChIP-exo analysis, including dPeak \cite{dpeak} and GEM \cite{gem}. We observe that dPeak performs as good or better than the available ChIP-exo methods and provides a platform where PE and SE ChIP-seq can be compared with their ChIP-exo counterpart. Our comparisons of PE ChIP-seq with ChIP-exo interestingly highlights that while ChIP-exo outperforms PE ChIP-seq in terms of resolution and detection power, both are significantly better than SE ChIP-seq.

% while there are not established QC pipelines for ChIP-exo; previous ChIP-exo analyses used ChIP-Seq samples to compare the resolution  between experiments \cite{exo1,exoillumina,exo2}.

%to interrogate these biases in a ChIP-exo experiment and globally asses the enrichment and library complexity of a ChIP-exo sample and a procedure to distinguish low complexity libraries from deeply sequenced experiments. This aspect is unique to ChIP-exo, since the exonuclease enzyme it is expected to digested the reads that are not bound to a transcription factor, therefore the number of bases where a ChIP-exo fragment could be potentially aligned is reduced. That way, in a high quality ChIP-exo experiment is possible to observe a large amount of reads to be aligned to unique position due to genomic signal instead of PCR artifacts.

%In order to obtain the potential benefits of ChIP-exo on protein binding site identification, it is critical to use algorithms that could fully utilize information available in ChIP-exo data. Rhee and Pugh, 2011 \nocite{exo1} discussed that reads in the forward and reverse strand might construct peak pairs around bound proteins, of which heights were implicitly assumed to be symmetric. Based on this rationale, they used the ``peak pair method'' that predicts the midpoint of two modes of peak pairs as potential binding sites. Recently developed ChIP-exo data analysis methods, such as Mace (Wang et al, 20114\nocite{mace}), CexoR (Madrigal, 2015\nocite{cexor}) and Peakzilla (Bardet et al., 2013\nocite{peakzilla}), are also based on this peak pair assumption. However, appropriateness of such assumption was not fully evaluated in the literature yet. Furthermore, it is still unknown which factors could affect protein binding site identification using ChIP-exo data. In order to address this problem, we investigated various aspects of ChIP-exo data by contrasting them with their respective ChIP-Seq experiments.

%Currently, research on statistical methods for ChIP-exo data is still in its very early stage. Although many methods have been proposed to %identify protein binding sites from ChIP-Seq data (reviewed by Wilbanks and Facciotti, 2012\nocite{evaluation} and Pepke and Wold, 2009 \nocite{computation}), such as MACS (Zhang et al., 2008\nocite{macs}), CisGenome (Ji et al., 2008\nocite{cisgenome}) and MOSAiCS (Kuan et al., 2009\nocite{mosaics}), these approaches might not fully utilize potentials of ChIP-exo data for high resolution identification of protein binding sites. Specifically these approaches reveal protein binding sites only in lower resolution, i.e., at an interval of hundreds to thousands of base pairs. Furthermore, they implicitly assume that there is only one ``mode'' or ``predicted binding location'' per this wide genomic interval. More recently, deconvolution algorithms such as Deconvolution (Lun et al., 2009\cite{csdeconv}), GEM (Guo et al., 2012\nocite{gem}, an improved version of Guo et al., 2010\nocite{gps} ) and PICS (Zhang et al., 2010\nocite{pics}) have been proposed to identify binding sites in higher resolution using ChIP-Seq data. However, most of them are still not tailored for ChIP-exo and PE and SE ChIP-Seq data in a unified framework and as a result, currently available methods are not appropriate for fair comparison between ChIP-exo and ChIP-Seq. To address these limitations, we developed and utilized an improved version of dPeak (Chung et al., 2013\nocite{dpeak}), a high resolution binding site identification (deconvolution) algorithm that we previously developed for PE and SE ChIP-Seq data, so that it can also handle ChIP-exo data. The dPeak algorithm implements a probabilistic model that accurately describes the ChIP-exo and ChIP-Seq data generation process.

%Some of the key findings in this work are as follows. First, we demonstrate that the ``peak pair'' assumption of Rhee and Pugh, 2013\nocite{exo2} does not hold well in real ChIP-exo data. Second, we found that when analyzing ChIP-exo data and the Input control is not available, it is useful to adjust for GC content and mappability biases to improve peak calling and binding site identification. Third, we evaluated several methods to identify binding events and dPeak performs competitively respect to GEM and MACE when analyzing ChIP-exo data. Finally, when comparable number of reads is used for both ChIP-exo and ChIP-Seq , dPeak coupled with ChIP-exo data provides resolution comparable to PE ChIP-Seq and both significantly improve the resolution of protein binding identification compared to SE-based analysis with any of the available methods.

@@

\subsection*{Publicly available ChIP-exo/nexus and novel \textit{E. Coli} ChIP-seq/exo datasets}

We utilized a rich collection of publicly available ChIP-exo/nexus data from multiple organisms to build and evaluate our quality control pipeline (Table~\ref{tab:qc}). These include: CTCF factor in human HeLa cells \cite{exo1}; ER factor in human MCF-7 cells \cite{exoillumina}; GR factor in IMR90, K562, and U2OS human cells \cite{starick15}; TBP factor in human K562 cells \cite{venters13}. ChIP-nexus data included experiments from \cite{chipnexus} profiling TBP in human K562 cells, MyC and Max in \textit{D. melanogaster} S2 cells, and Twist and Dorsal in \textit{D. melanogaster} embryo.

In order to have a setting where we can compare SE and PE ChIP-seq with their ChIP-exo counterpart, we profiled $\sig$ under a variety of conditions in \textit{E. coli} with both ChIP-exo and PE ChIP-seq (Table~\ref{tab:qc_sig}). Collectively, we generated $\sig$ factor ChIP-exo, PE and SE ChIP-seq experiments under aerobic ($+O_2$) and anaerobic ($-O_2$) conditions in glucose minimal media. For simplicity, we named these experiments as E1, P1, and S1, respectively. Similarly, we generated $\sig$ factor ChIP-exo, PE and SE (generated \textit{in   silico}) ChIP-seq experiments in \textit{E. Coli} under aerobic ($+O_2$) conditions with and without rifampicin treatment. We refer to these latter set of experiments as time point 20 minutes and time point 0 minutes. We also named these experiments E2, P2, and S2 for ChIP-exo, PE ChIP-seq and SE ChIP-seq, respectively. We specifically used these experimental data for comparisons of ChIP-exo and PE ChIP-seq assays in identifying closely spaced binding events and their resolution, i.e., physical proximity of the predicted events to the actual binding sites.  This comparison benefits from using $\sig$ which is a transcription initiation factor of housekeeping genes in \textit{E. coli}. Many \textit{E. Coli} promoters contain multiple transcription start sites (TSS). These TSSs are often closely spaced, i.e., within 10 $\sim$ 150 bps of each other, and are considered to be multiple ``switches'' that differentially regulate gene expression under diverse growth conditions \cite{regulondb}.

% \SK{Check for consistency with other parts of the paper and correct, ok to use different naming convention}.

@@

@@

\subsubsection*{Saturation analysis with ChIP-exo and ChIP-seq: Systematic comparison of ChIP-seq vs. ChIP-exo under
  varying sequencing depths}
Establishing competitive performance of dPeak compared to other
ChIP-exo analysis methods enable comparison of ChIP-exo and ChIP-seq
with a unified analysis framework using dPeak. Previous comparisons of
ChIP-exo and SE ChIP-seq were all performed without controlling for
the sequencing depths of the samples. More importantly, although it is
well established that PE ChIP-seq leads to better resolution that SE
ChIP-seq in terms of binding site identification \cite{zhang16},
previous work does not have an in depth comparison of ChIP-exo and PE
ChIP-seq.
%Previously, ChIP-exo and SE ChIP-Seq have been compared only at a
%fixed depth level in the literature, while they did not include PE
%ChIP-Seq as well either. 
In order to address these limitations of the previous studies, we
performed a sub-sampling experiment by sampling fixed numbers of reads
from each of the $\sig$ ChIP-exo, PE ChIP-seq, and SE ChIP-seq
datasets. Specifically, for every $N$ reads in ChIP-exo and SE
ChIP-seq, $N/2$ read pairs were sampled for PE ChIP-seq to operate
under fixed sequencing costs.
%Additionally, it is worth noticing that for PE ChIP-Seq we
%sampled both ends of the fragment, hence for each sequencing depth we
%are sampling the half amount of pairs for PE ChIP-Seq than for
%ChIP-exo or SE ChIP-Seq.
